---
title: "MBA Data Exploration"
author: "Scott Bailey"
date: "4/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse, warn.conflicts = FALSE)
library(tidylog, warn.conflicts = FALSE)
library(skimr)
library(mgcv)
library(lubridate)
library(gratia)
library(forcats)
```

```{r data-imports}
features <- read_csv("data/features.csv")
stores <- read_csv("data/stores.csv")
test_df <-read_csv("data/test.csv")
train_df <- read_csv("data/train.csv")
sample_submission_df <- read_csv("data/sampleSubmission.csv")
```
```{r}
skim(train_df)
```
```{r}
train_df %>%
  filter(Store == 1) %>%
  filter(Dept == 1) %>%
  ggplot(aes(Date, Weekly_Sales, color = IsHoliday)) +
  geom_point()
```

Let's try to model a single department of a single store using a generalized additive model. We'll need to break apart the Date column into numeric variables
 
```{r}
train_dat_1_1 <- train_df %>%
  filter(Store == 1) %>%
  filter(Dept == 1) %>%
  mutate(
    Day = day(Date),
    Month = month(Date),
    Year = year(Date),
    Week = week(Date)
  )

head(train_dat_1_1)
```
 
```{r}
ggplot(train_dat_1_1, aes(x=IsHoliday, y=Weekly_Sales)) +
  geom_point()
```
 
 
```{r}
gam_mod <- gam(Weekly_Sales ~ s(Week), data = train_dat_1_1, method="REML")

draw(gam_mod) 
plot(gam_mod, residuals = TRUE, pch = 1, cex = 1, shade = TRUE)
```
```{r}
ggplot(data = train_dat_1_1, aes(x = Week, y = Weekly_Sales, color = Year)) +
  geom_point(aes(shape = IsHoliday)) +
  geom_line(aes(x=Week, y=fitted(gam_mod)))
```
 
 Let's take a look at the full training data. 
 
```{r}
train_mod <- train_df %>%
  mutate(
    Day = day(Date),
    Month = month(Date),
    Year = year(Date),
    Week = week(Date),
    Store = as_factor(Store),
    Dept = as_factor(Dept)
  )
  

head(train_mod)
```
 
Let's do some exploratory graphing with some groupby statements. 

```{r}
train_mod %>%
  group_by(Store) %>%
  summarise(total_sales = sum(Weekly_Sales)) %>%
  ggplot(aes(x = Store, y = total_sales)) +
  geom_col()
  
  
```
```{r}
train_mod %>%
  group_by(Dept) %>%
  summarise(total_sales = sum(Weekly_Sales)) %>%
  ggplot(aes(x = Dept, y = total_sales)) +
  geom_col()
```

```{r}
train_mod %>%
  ggplot(aes(x = Week, y = Weekly_Sales, color = Year)) +
  geom_point(aes(shape = IsHoliday))
```


Train a GAM on whole data - THESE ARE SLOW AND NOT EFFECTIVE BC THE DATA DOESNT HAVE AN UNDERLYING RELATIONSHIP. THEY HAVE TO BE GROUPED BY STORE
```{r}
gam_mod_w <- gam(Weekly_Sales ~ s(Week) + Store + Dept, data = train_mod, method="REML")

gam_mod_w <- gam(Weekly_Sales ~ s(Week, by=Store) + Store, data = train_mod, method="REML")
```
```{r}
train_1 <- train_mod %>%
  filter(Store == 1)

train_1 %>%
  ggplot(aes(x = Week, y = Weekly_Sales)) +
  geom_point() +
  facet_wrap(vars(Dept))
```

 
 Need to group by store to get aggregate sum of weekly sales
 Facet on store, department, year
